---
title: "Comparative effect of thrombectomy"
format: gfm
embed-resources: true
editor: visual
execute: 
  warning: false
  error: false
  cache: false
editor_options: 
  chunk_output_type: console
---

```{r load packages, data, convenience functions}
#| echo: false
library(here)
library(tidyverse)
library(stringr)

data <- read_csv(here("data", "clean_data.csv"), show_col_types = FALSE)

varying_summary <- readRDS(here("fits", "varying_summary.RDS")) 

comma <- function(x) format(x, digits = 3, big.mark = ",", small.mark = ",")
```

## Observed Data

```{r calculations}
#| echo: false

trials <- data %>% 
  nrow()
patients_total <- with(data, sum(n_c) + sum(n_t)) %>% 
  comma()
patients_control <- with(data, sum(n_c)) %>%
  comma()
patients_treatment <- with(data, sum(n_t)) %>% 
  comma()
categories <- n_distinct(data$K)
```

There were a total of N = `r patients_total` patients enrolled in J = `r trials` trials. `r patients_control` were randomized to best medical management. `r patients_treatment` were randomized to best medical management plus mechanical thrombectomy. Each trial was categorized into one of K = `r categories` categories according to the type of stroke patients enrolled. Category 1 was large anterior circulation strokes. Category 2 was small to medium sized anterior circulation strokes treated in an early time window. Category 3 was small to medium sized anterior circulation strokes treated in a late time window. Category 4 was strokes involving the basilar artery. The primary endpoint was functional independence at 90 days (modified Rankin score of 0 to 2).

Results from the large stroke trials are shown here, with $y$ representing the observed treatment effect of thrombectomy (the odds ratio on the log scale) and $\text{se}$ representing the standard error for $y$.

```{r}
#| echo: false
data <- data %>% 
  mutate(name = c("ANGEL", "RESCUE", "SELECT2", "TENSION", "TESLA",
                   "ESCAPE", "EXTEND", "MRCLEAN", "PISTE", "RESILIENT", "REVASCAT", "SWIFT", "THERAPY", "THRACE",
                   "DAWN", "DEFUSE3", "MRCLEANLATE", "POSITIVE",
                   "ATTENTION", "BAOCHE", "BASICS", "BEST")) %>% 
  relocate("name", .after = J)

data %>% filter(K == 1)
```

Results from the early stroke trials are shown here.

```{r}
#| echo: false
data %>% filter(K == 2)
```

Results of the late stroke trials are shown here.

```{r}
#| echo: false
data %>% filter(K == 3)
```

Basilar stroke trial results are shown here.

```{r}
#| echo: false
data %>% filter(K == 4)
```

The pooled data by category are shown in the following table.

```{r pooled or table}
#| echo: false
data %>% 
  group_by(K) %>% 
  summarise(trials = n(),
            "sample size" = sum(n_c + n_t),
            r_c = sum(r_c),
            n_c = sum(n_c),
            r_t = sum(r_t),
            n_t = sum(n_t)) %>% 
  mutate(
    y = log(r_t / (n_t - r_t)) - log(r_c / (n_c - r_c)),
    se = sqrt(1/r_t + 1/(n_t - r_t) + 1/r_c + 1/(n_c - r_c)),
    category = c("large", "early", "late", "basilar")
  ) %>% 
  select(category, trials, "sample size", y, se )
```

The overall pooled data is shown in the following table.

```{r pooled or}
#| echo: false
(
  pooled <- data %>% 
  summarise(trials = n(),
            "sample size" = sum(n_c + n_t),
            r_c = sum(r_c),
            n_c = sum(n_c),
            r_t = sum(r_t),
            n_t = sum(n_t)) %>% 
  mutate(
    y = 
      log(r_t / (n_t - r_t)) - log(r_c / (n_c - r_c)),
    se = 
      sqrt(1/r_t + 1/(n_t - r_t) + 1/r_c + 1/(n_c - r_c))
  ) %>% 
    select(trials, "sample size", y, se )
  )
```

## Estimated average treatment effect of thrombectomy across trials

### Early

The model estimate for the average treatment effect of thrombectomy across early stroke trials is summarized in the following table.

```{r}
#| echo: false
(
  df <- varying_summary %>% 
  filter(model == "varying",
         data == "early",
         priors == "diffuse",
         variable == "mu") %>% 
  select("median", "mad", "low", "high", "prob_pos")
)
```

These values correspond to an estimated `r comma((exp(df$median) - 1)*100)`% average improvement in the odds of functional independence at 90 days with thrombectomy (95% CI `r comma((exp(df$low) - 1)*100)`% to `r comma((exp(df$high) - 1)*100)`%, P(+) = `r comma(df$prob_pos*100)`%).

### Large

The model estimate for the average treatment effect of thrombectomy across large stroke trials is summarized in the following table.

```{r}
#| echo: false
(df <- varying_summary %>% 
  filter(model == "varying",
         data == "large",
         priors == "diffuse",
         variable == "mu") %>% 
  select("median", "mad", "low", "high", "prob_pos")
)
```

These values correspond to an estimated `r comma((exp(df$median) - 1)*100)`% average improvement in the odds of functional independence at 90 days with thrombectomy (95% CI `r comma((exp(df$low) - 1)*100)`% to `r comma((exp(df$high) - 1)*100)`%, P(+) = `r comma(df$prob_pos*100)`%).

### Late

The model estimate for the average treatment effect of thrombectomy across late stroke trials is summarized in the following table.

```{r}
#| echo: false
(df <- varying_summary %>% 
  filter(model == "varying",
         data == "late",
         priors == "diffuse",
         variable == "mu") %>% 
  select("median", "mad", "low", "high", "prob_pos")
)
```

These values correspond to an estimated `r comma((exp(df$median) - 1)*100)`% average improvement in the odds of functional independence at 90 days with thrombectomy (95% CI `r comma((exp(df$low) - 1)*100)`% to `r comma((exp(df$high) - 1)*100)`%, P(+) = `r comma(df$prob_pos*100)`%).

### Basilar

The model estimate for the average relative treatment effect of thrombectomy across basilar stroke trials is summarized in the following table.

```{r}
#| echo: false
(df <- varying_summary %>% 
  filter(model == "varying",
         data == "basilar",
         priors == "diffuse",
         variable == "mu") %>% 
  select("median", "mad", "low", "high", "prob_pos")
)
```

These values correspond to an estimated `r comma((exp(df$median) - 1)*100)`% average improvement in the odds of functional independence at 90 days with thrombectomy (95% CI `r comma((exp(df$low) - 1)*100)`% to `r comma((exp(df$high) - 1)*100)`%, P(+) = `r comma(df$prob_pos*100)`%).

## Anticipated relative treatment effect of thrombectomy in a new trial

The model estimate for the variation in the trial-specific treatment effect of thrombectomy for each stroke type is summarized in the following table.

```{r}
#| echo: false
(
  df <- varying_summary %>% 
  filter(model == "varying",
         data %in% c("early", "large", "late", "basilar"),
         priors == "diffuse",
         variable == "sigma") %>% 
  select("data", "median", "mad", "low", "high")
)
```

When taken together with the estimates for the average relative treatment effects, these estimates of trial-specific treatment effect heterogeneity can be used to forecast the expected treatment effect of thrombectomy in a new trial, which we call the anticipated treatment effect.

```{r}
#| echo: false
(
  df <- varying_summary %>% 
    filter(model == "varying",
           data %in% c("early", "large", "late", "basilar"),
           priors == "diffuse",
           variable == "theta_new") %>% 
    select("data", "median", "mad", "low", "high", "prob_pos")
)
```

### Early

```{r}
#| echo: false
df <- varying_summary %>% 
  filter(model == "varying",
         data == "early",
         priors == "diffuse",
         variable == "theta_new") %>% 
  select("data", "median", "mad", "low", "high", "prob_pos")
```

The anticipated treatment effect of thrombectomy in a new early stroke trial has an odds ratio of `r comma(exp(df$median))` (95% CI `r comma(exp(df$low))` to `r comma(exp(df$high))`, P(+) = `r comma(df$prob_pos*100)`%).

### Large

```{r}
#| echo: false
df <- varying_summary %>% 
  filter(model == "varying",
         data == "large",
         priors == "diffuse",
         variable == "theta_new") %>% 
  select("data", "median", "mad", "low", "high", "prob_pos")
```

The anticipated treatment effect of thrombectomy in a new large stroke trial has an odds ratio of `r comma(exp(df$median))` (95% CI `r comma(exp(df$low))` to `r comma(exp(df$high))`, P(+) = `r comma(df$prob_pos*100)`%).

### Late

```{r}
#| echo: false
df <- varying_summary %>% 
  filter(model == "varying",
         data == "late",
         priors == "diffuse",
         variable == "theta_new") %>% 
  select("data", "median", "mad", "low", "high", "prob_pos")
```

The anticipated treatment effect of thrombectomy in a new late stroke trial has an odds ratio of `r comma(exp(df$median))` (95% CI `r comma(exp(df$low))` to `r comma(exp(df$high))`, P(+) = `r comma(df$prob_pos*100)`%).

### Basilar

```{r}
#| echo: false
df <- varying_summary %>% 
  filter(model == "varying",
         data == "basilar",
         priors == "diffuse",
         variable == "theta_new") %>% 
  select("data", "median", "mad", "low", "high", "prob_pos")
```

The anticipated treatment effect of thrombectomy in a new basilar stroke trial has an odds ratio of `r comma(exp(df$median))` (95% CI `r comma(exp(df$low))` to `r comma(exp(df$high))`, P(+) = `r comma(df$prob_pos*100)`%).
